Investigating magnetic dynamo evolution with TESS field dwarfs

### 2017-10-05 ms.tex
editing on the train

In addition, we will use the rotation periods of comoving stars identified in
the first \Gaia\ data release \citep{oh2016} to quantify the accuracy and
precision of gyrochronology.
We have identified 38 stars in 19 pairs with \Gaia\ proper motions and radial
velocities from \RAVE\ that are observable by \TESS, using {\tt tvguide}.
Of these, 34 stars in 17 pairs are in the CTL list.
Since these stars would make interesting targets for both asteroseismology and
exoplanet search, we include a list of them with this proposal.
The stars in these pairs are assumed to be recently disrupted binaries so we
expect them to have the same age; measuring their rotation periods will
therefore provide a test of gyrochronology.

\paragraph{Technical Feasibility}
The extraction of a rotation period from a light curve can be as simple as
computing a Lomb-Scargle periodogram or, in cases where the signal is less
clear, can be inferred via modeling the correlation between data points.
This latter approach could involve either computing an autocorrelation
function \citep{mcquillan2013}, or fitting a Gaussian process to the time
series \citep{angus2017, foreman-mackey2017}.
In cases where signals have long periods and low amplitudes, care is needed to
separate real astrophysical signal from instrumental systematics.
The measurement of rotation periods is less sensitive to crowding and source
confusion than exoplanet transit characterization because the rotation period
is not effected by photometric dilution.
We will apply two complementary methods to extract and calibrate light curves
from the TESS FFIs.
First, for bright or isolated targets, we will follow \citet{montet2017} to
estimate aperture shapes and perform aperture photometry for bright sources.
Using an ensemble of sources, we will de-trend these light curves using a
modified version of \textsf{everest} \citep{luger2016, luger2017} designed to
preserve photometric signatures of rotation.
This will be achieved by fitting for the systematic effects in the light curve
using the \textsf{everest} model simultaneously with a Gaussian Process model
for the astrophysical variation.
Both \citet{aigrain2016} and \citet{luger2016} demonstrated that this can
preserve stellar variability signals and we will use the \textsf{celerite}
algorithm \citep{dfm2017} to scale the computations to the size of TESS FFI
datasets.
The precision of existing photometric de-trending methods degrades in crowded
fields \citep[for example,][]{luger2017}.
However, to make robust measurements of rotation periods, we do not need
absolutely calibrated light curves.
Therefore, in crowded fields, we will apply a difference imaging method that
was developed for the K2 Campaign 9 microlensing project \citep{henderson2016}
based on the \textsf{CPM} \citep{wang2016} to robustly measure the photometric
variations of crowded sources.
Unlike standard difference imaging methods, this procedure does not require a
reference image.
Instead, a causal data-driven model is built to predict the time series in
every pixel taking systematic effects into account and the residuals between
the observations and the model predictions provide an estimate of the
astrophysical variability in each pixel.
We will tune this method preserve rotation signals and apply it to detect
rotation across the FFIs.

The full set of available \TESS\ FFI light curves will be exceedingly large
and we intend to prioritize certain groups of stars for period analysis,
calculating a simple (and computationally fast) Lomb-Scargle periodogram for
the low priority objects and running a full probabilistic period analysis on
the high priority targets.
High priority targets include: comoving pairs, stars in the CVZs (especially
those with spectra), stars in TGAS, stars at large distances, plus any
potential asteroseismic dwarfs and exoplanet hosts.
Based on the expected photometric precision of \tess\ \citep{sullivan2015}, we
expect to be able to measure rotation periods for FFI stars down to 16th
magnitude.
We note, however that not all stars will have measurable rotation periods,
even if they are bright enough --- not all stars show rotational variability
in their light curves either because they are pole-on or because they are
inactive.

@@

\paragraph{Expected Impact}
Initially, we will provide rotation period posterior probability density
functions and best fit estimates for around 30,000 high priority FFI targets.
We will then produce a catalog of rotation periods for lower priority FFI
stars using a combination of Lomb-Scargle periodogram and autocorrelation
techniques.
Finally, we will release a catalog of rotation periods for two-minute cadence
targets.
The full catalog will likely contain several million rotation periods.
We will also release an open source python package for measuring rotation
periods of both FFI and two-minute cadence targets.

\paragraph{Budget Justification}
PI Angus intends to use the budget to employ a student for 3-4 months.
In this time the student will assist in extracting light curves from the
\tess\ FFIs, measuring rotation periods and building a rotation period
catalog.
The student will also be involved in the scientific project of their choosing:
either the rotation period bimodality, gyrochronology of low-mass stars, or
gyrochronology of comoving pairs.
The remaining budget will be used for traveling to meet with other Co-Is in
order to perform this analysis.
\paragraph{References}

>>>>

In addition, we will use the rotation periods of comoving stars identified in
the first \Gaia\ data release \citep{oh2016} to quantify the accuracy and
precision of gyrochronology.
We have identified 38 stars in 19 pairs with \Gaia\ proper motions and radial
velocities from \RAVE\ that are observable by \TESS, using {\tt tvguide}.
Of these, 34 stars in 17 pairs are in the CTL list.
Since these stars would make interesting targets for both asteroseismology and
exoplanet search, we include a list of them with this proposal.
The stars in these pairs are likely recently disrupted binaries so we
expect them to have the same age and measuring their rotation periods will
therefore provide a test of gyrochronology.

\paragraph{Technical Feasibility}
The extraction of a rotation period from a light curve can be as simple as
computing a Lomb-Scargle periodogram or, in cases where the signal is less
clear, can be inferred via modeling the correlation between data points.
This latter approach could involve either computing an autocorrelation
function \citep[\eg][]{mcquillan2013}, or fitting a Gaussian process to the
time series \citep{angus2017, foreman-mackey2017}.
In cases where signals have long periods and low amplitudes, care is needed to
separate real astrophysical signal from instrumental systematics.
The measurement of rotation periods is less sensitive to crowding and source
confusion than exoplanet transit characterization because the rotation period
is not effected by photometric dilution.
We will apply two complementary methods to extract and calibrate light curves
from the TESS FFIs.
First, for bright or isolated targets, we will follow \citet{montet2017} to
estimate aperture shapes and perform aperture photometry for bright sources.
Using an ensemble of sources, we will de-trend these light curves using a
modified version of \textsf{everest} \citep{luger2016, luger2017} designed to
preserve photometric signatures of rotation.
This will be achieved by fitting for the systematic effects in the light curve
using the \textsf{everest} model simultaneously with a Gaussian process model
for the astrophysical variation.
Both \citet{aigrain2016} and \citet{luger2016} demonstrated that this can
preserve stellar variability signals and we will use the \textsf{celerite}
algorithm \citep{foreman-mackey2017} to scale the computations to the size of
TESS FFI datasets.
The precision of existing photometric de-trending methods degrades in crowded
fields \citep[for example,][]{luger2017}.
However, to make robust measurements of rotation periods, we do not need
absolutely calibrated light curves.
Therefore, in crowded fields, we will apply a difference imaging method that
was developed for the K2 Campaign 9 microlensing project \citep{henderson2016}
based on the \textsf{CPM} \citep{wang2016} to robustly measure the photometric
variations of crowded sources.
Unlike standard difference imaging methods, this procedure does not require a
reference image.
Instead, a causal data-driven model is built to predict the time series in
every pixel taking systematic effects into account and the residuals between
the observations and the model predictions provide an estimate of the
astrophysical variability in each pixel.
We will tune this method to preserve rotation signals and apply it to detect
rotation across the FFIs.

The full set of available \TESS\ FFI light curves will be exceedingly large
and we intend to prioritize certain groups of stars for period analysis,
calculating a simple (and computationally fast) Lomb-Scargle periodogram for
the low priority objects and running a full probabilistic, Gaussian process
period analysis on the high priority targets.
High priority targets include: comoving pairs, stars in the CVZs (especially
those with spectra), stars in TGAS, stars at large distances, plus any
potential asteroseismic dwarfs and exoplanet hosts.
Based on the expected photometric precision of \tess\ \citep{sullivan2015}, we
expect to be able to measure rotation periods for FFI stars down to 16th
magnitude.
We note, however that not all stars will have measurable rotation periods,
even if they are bright enough --- not all stars show rotational variability
in their light curves either because they are pole-on or because they are
inactive.

@@

\paragraph{Expected Impact}
Initially, we will provide rotation period posterior probability density
functions and best fit estimates for around 30,000 high priority FFI targets.
We will then produce a catalog of rotation periods for lower priority FFI
stars using a combination of Lomb-Scargle periodogram and autocorrelation
techniques.
Finally, we will release a catalog of rotation periods for all two-minute
cadence targets.
The full catalog will likely contain several million rotation periods.
We will release code for inferring stellar ages from rotation periods, based
on a new gyrochronology relation calibrated using \TESS\HERMES spectra, \Gaia\
kinematics and comoving pairs.
We will also release an open source python package for measuring rotation
periods of both FFI and two-minute cadence targets.

\paragraph{Budget Justification}
PI Angus intends to use the budget to employ an undergraduate student for 3-4
months.
In this time the student will assist in extracting light curves from the
\tess\ FFIs, measuring rotation periods and building a rotation period
catalog.
The student will also be involved in the scientific project of their choosing:
either the rotation period bimodality, gyrochronology of low-mass stars, or
gyrochronology of comoving pairs.
The remaining budget will be used for traveling to meet with other Co-Is in
order to perform this analysis.
\paragraph{References}
